Affective Uplift During Video Game Play: A Naturalistic Case Study

### 2023-10-18 ms.Rmd
Notebooks to reproducible manuscript.

---
title: ''
shorttitle: 
leftheader: 
author: 
  - name: Matti Vuorre
    affiliation: 1
    corresponding: yes
    address: Department of Social Psychology, Tilburg University
    email: m.j.vuorre@tilburguniversity.edu
affiliation:
  - id: 1
    institution: Department of Social Psychology, Tilburg University
abstract: |

keywords: 
authornote: |
  \noindent \textbf{Working paper. Proceed with caution.}
wordcount: "`r wordcountaddin:::word_count(filename = 'ms.Rmd')`"
bibliography: references.bib
floatsintext: yes
numbersections: false
linenumbers: no
draft: no
mask: no
figurelist: no
tablelist: no
footnotelist: no
documentclass: apa7
classoption: jou
header-includes:
  - |
    \usepackage{float}
    \setlength{\parskip}{4pt}
output: 
  papaja::apa6_pdf:
    number_sections: true
    keep_tex: false
    highlight: kate
  papaja::apa6_docx: 
    number_sections: true
    keep_tex: false
---

```{r setup, include = FALSE}
# Packages
library(tidyverse)

# Output options
opts_chunk$set(
  eval = TRUE,
  cache = TRUE,
  warning = FALSE,
  error = TRUE,
  message = FALSE
)

# Plotting options
theme_set(
  theme_linedraw(base_size = 12) +
    theme(
      strip.background = element_rect(fill = "gray90", color = "gray90"),
      strip.text = element_text(color = "black", hjust = 0),
      panel.grid = element_blank()
    )
)

```

# References {-}

::: {#refs custom-style="Bibliography"}
:::

>>>>

---
title: "Affective Uplift During Video Game Play: A Naturalistic Case Study"
shorttitle: Affective Uplift During Play
leftheader: Affective Uplift During Play
author: 
  - name: Matti Vuorre
    corresponding: yes
    affiliation: "1,2*"
    address: Tilburg University
    email: mjvuorre@uvt.nl
  - name: Nick Ballou
    affiliation: "2*"
  - name: Thomas Hakman
    affiliation: "2"
  - name: Kristoffer Magnusson
    affiliation: "2,3"
  - name: Andrew K. Przybylski
    affiliation: "2"
    address: Oxford Internet Institute
    email: andy.przybylski@oii.ox.ac.uk
affiliation:
  - id: 1
    institution: Tilburg School of Social and Behavioral Sciences, Tilburg University
  - id: 2
    institution: Oxford Internet Institute, University of Oxford
  - id: 3
    institution: Centre for Psychiatry Research, Department of Clinical Neuroscience, Karolinska Institutet, & Stockholm Health Care Services, Region Stockholm, Sweden
authornote: |
  \noindent \textbf{This pre-print is not yet peer-reviewed.} 
  *MV and NB contributed equally to this work.

wordcount: '`r scales::number(wordcountaddin:::word_count("ms.Rmd"), big.mark = ",")`'
bibliography: references.bib
keywords: "video games, well-being, mood, play behavior, telemetry"
csl: apa.csl
floatsintext: yes
linenumbers: no
draft: no
mask: no
figurelist: no
tablelist: no
footnotelist: no
documentclass: apa7
classoption: jou
output: 
  papaja::apa6_pdf:
    number_sections: false
    keep_tex: true
  papaja::apa6_docx: 
    number_sections: false
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: prepare
#| cache: false

library(scales)
library(cmdstanr)
library(splines)
library(emmeans)
library(posterior)
library(lme4)
library(brms)
library(gt)
library(ggstance)
library(ggdist)
library(patchwork)
library(tidyverse)
source("R/functions.R")
source("R/common.R")

dir.create("models", FALSE)
```

# Introduction

How do video games affect players' well-being? Games are often studied for their potential in catalyzing psychological change over timescales spanning from weeks to months [e.g., effects on school performance, depression, or life satisfaction, @sauterSocialContextGaming2021; @vuorreTimeSpentPlaying2022], and the surrounding public debate has typically focused on play's far-reaching consequences on players' mental health, social attitudes, or cognitive development [@fergusonDoesSexualizationVideo2022; @hilgardOverestimationActionGameTraining2019; @mathurFindingCommonGround2019]. In stark contrast, typical play appears to be motivated by short-term goals, such as wanting to unwind after a long day, escape to a pleasant non-reality in the moment, or engage in uplifting social interaction over periods of hours [@bourgonjonPlayersPerspectivesPositive2016; @kahnTrojanPlayerTypology2015; @stensengAreThereTwo2021]. Such short-term dynamics between play and affect can exist but need not necessarily accumulate into long-term impacts. For example, games might provide relief, relaxation, and brief improvements in mood over several hours [@riegerEatingGhostsUnderlying2015; @russonielloEffectivenessCasualVideo2009; @tyackRestorativePlayVideogames2020], after which the effects taper out as individuals return to their baseline moods.

Understanding whether and when games' short-term effects emerge is critical for establishing games' potential for mood-related interventions, as well as for building a theoretical foundation for repeated short-term gaming experiences' long-term effects on mental health. At present, however, the prevalence and magnitude of these short-term effects remains poorly understood. Some existing evidence suggests that games can provide short-term boosts to well-being ["mood repair"; @bowmanMoodGameSelective2015; @tyackRestorativePlayVideogames2020], possibly to a greater extent than non-interactive media such as videos [@riegerEatingGhostsUnderlying2015]. On the other hand, games might affect players negatively: Frustrating gaming experiences, for example, can lead to negative consequences such as immediate post-play aggression [@przybylskiCompetenceimpedingElectronicGames2014].

Despite the above examples, the validity and generalizability of research on games' short-term affective effects remains limited by three challenges. First, a substantial portion of gameplay research has relied on artificial stimuli; games created or substantially modified by academic researchers. While such customized games allow for greater experimental control, they are unlikely to reflect actual games' rich complexity [@mcmahanConsiderationsUseCommercial2011]. This issue of the limited ecological validity and generalizability of research stimuli (games) limits current inferences about popularly played games' psychological effects.

The second challenge is providing an ecologically valid *context* for play. Research participants typically play games in (online or physical) labs that do not resemble the natural contexts of play, such as when, with whom, and why people choose to play. In lab settings, research participants play to satisfy study requirements, rather than the intrinsic motivations that typically lead them to play. While beneficial to clarifying causal inferences, the extrinsically motivated play behaviors necessary in lab studies might relate differently to well-being than intrinsically motivated naturally occurring play [@bruhlmannMotivationalProfilingLeague2020; @howardStudentMotivationAssociated2021]. Therefore, results from such studies are less likely to accurately generalize to how games are played in the real world.

The third challenge concerns the timescale of effects: How quickly do potential effects emerge, and how long are they sustained? For example, some studies indicate that by the end of a half-hour game session, players may exhibit changes in stress [@russonielloEffectivenessCasualVideo2009], aggressive affect [@przybylskiCompetenceimpedingElectronicGames2014], and vitality [@tyackRestorativePlayVideogames2020]. When and how video games' effects evolve during the initial half-hour remains unclear and difficult to study because researchers are typically unable to ask questions at a sufficient temporal resolution, with notable exceptions of @boweyPredictingBeliefsNPC2021 and @frommelGatheringSelfReportData2021, who used non-player characters to ask questions directly within the game. However, they did not enquire about well-being, leaving it unclear when and how the affective dimensions of play change on short timescales.

Here, we aim to address these three challenges to better understand how real play in natural contexts might predict mood on short timescales. Specifically, we examined an intensive longitudinal dataset from the popular commercially available game PowerWash Simulator [PWS; @vuorreIntensiveLongitudinalDataset2023], which includes mood questions embedded in the game itself, to ask three questions: First, to what extent does mood change from immediately before video game play to during play? Second, hyow heterogeneous are these changes in the population of similar players? And third, how do changes in mood develop over the course of a gaming session?

## Methods

<!-- Data wrangling code -->

```{r}
#| label: data-get

# Download, unzip data from OSF PWS database
path <- "data-raw/data.zip"
if (!file.exists(path)) {
  dir.create("data-raw", FALSE)
  download.file(
    url = "https://osf.io/download/j48qf/",
    destfile = "data-raw/data.zip"
  )
  dir.create("data", FALSE)
  unzip(
    zipfile = path,
    files = c(
      "data/demographics.csv",
      "data/study_prompt_answered.csv"
    )
  )
}

# Load data to R session
dat <- read_csv(
  "data/study_prompt_answered.csv",
  col_select = c(
    pid,
    time = Time_utc,
    duration = CurrentSessionLength,
    prompt = LastStudyPromptType,
    mood = response
  )
)

# Convert types and units
dat <- dat |>
  mutate(
    pid = factor(pid),
    # Mood to 0-1 scale
    mood = mood / 1000,
    # hours indicates session duration in hours
    hours = duration / 60,
    .keep = "unused"
  )
```

```{r}
#| label: session-indicator-create
#| include: true

dat <- dat |>
  arrange(pid, time) |>
  mutate(
    new_session = hours < lag(hours, default = 999),
    session = cumsum(new_session),
    .by = pid
  ) |>
  select(-c(new_session)) |>
  mutate(ps = paste(pid, session, sep = "_"))

dat_sum <- dat |>
  summarize(
    when = "raw",
    n_pid = length(unique(pid)),
    n_session = length(unique(ps)),
    n_obs = n()
  )
```

```{r}
#| label: mood-filter

# Include only wellbeing responses that occurred
# in sessions with both "pre" and "post" measures
dat <- dat |>
  filter(prompt == "Wellbeing") |>
  select(-prompt)

dat_sum <- dat |>
  summarize(
    when = "mood",
    n_pid = length(unique(pid)),
    n_session = length(unique(ps)),
    n_obs = n()
  ) |>
  bind_rows(dat_sum)
```

```{r}
#| label: fig-session-durations
#| include: false
#| fig-height: 3
#| fig.cap: Summaries of session durations. X-axes are truncated at 10 hours.

p_durations <- dat |>
  filter(hours == max(hours), .by = ps) |>
  ggplot(aes(hours)) +
  scale_x_continuous(
    "Session duration",
    expand = expansion(c(0.05, 0.05))
  ) +
  scale_y_continuous(
    "Sessions",
    expand = expansion(c(0, 0.05))
  ) +
  coord_cartesian(xlim = c(0, 10)) +
  geom_histogram(binwidth = 0.33)

p_mean_durations <- dat |>
  filter(hours == max(hours), .by = ps) |>
  summarise(hours = mean(hours), n = n(), .by = pid) |>
  ggplot(aes(hours)) +
  scale_x_continuous(
    "Mean session duration",
    expand = expansion(c(0.05, 0.05))
  ) +
  scale_y_continuous(
    "Players",
    expand = expansion(c(0, 0.05))
  ) +
  coord_cartesian(xlim = c(0, 10)) +
  geom_histogram(binwidth = 0.33)

p_ecdf_durations <- dat |>
  filter(hours == max(hours), .by = c(session, pid)) |>
  ggplot(aes(hours)) +
  scale_x_continuous(
    "Session duration",
    expand = expansion(c(0.05, 0.05))
  ) +
  scale_y_continuous(
    "Cumulative proportion",
    expand = expansion(0.01),
    breaks = (0:10) / 10
  ) +
  stat_ecdf() +
  coord_cartesian(xlim = c(0, 10))

p_durations | p_mean_durations | p_ecdf_durations

dat <- dat |>
  filter(hours <= 5)

dat_sum <- dat |>
  summarize(
    when = "5h",
    n_pid = length(unique(pid)),
    n_session = length(unique(ps)),
    n_obs = n()
  ) |>
  bind_rows(dat_sum)
```

```{r}
#| label: wrangle-more
#| include: false

# Drop rows with missing mood responses
dat <- dat |>
  drop_na(mood) |>
  mutate(pid = fct_drop(pid))

# Create indicators for analyses
dat <- dat |>
  mutate(
    post = factor(
      row_number() > 1,
      levels = c(FALSE, TRUE),
      labels = c("0", "1")
    ),
    .by = c(pid, session)
  ) |>
  # Censoring indicator
  mutate(
    cl = case_when(
      mood == 0 ~ "left",
      mood == 1 ~ "right",
      TRUE ~ "none"
    )
  )

dat_sum <- dat |>
  summarize(
    when = "non-na",
    n_pid = length(unique(pid)),
    n_session = length(unique(ps)),
    n_obs = n()
  ) |>
  bind_rows(dat_sum)

# How many sessions had pre-mood & post-mood
dat |>
  mutate(
    has_pre = any(hours == 0),
    has_post = any(hours != 0),
    has_both = has_pre & has_post,
    .by = ps
  ) |>
  distinct(ps, has_pre, has_post, has_both) |>
  summarise(
    n_pre = number2(sum(has_pre)),
    n_post = number2(sum(has_post)),
    n_both = number2(sum(has_both)),
    p_pre = percent2(mean(has_pre)),
    p_post = percent2(mean(has_post)),
    p_both = percent2(mean(has_both))
  )

# Use contrast coding with unit difference
contrasts(dat$post) <- c(-0.5, 0.5)

# Write data for any supplementary notebooks
dir.create("data", FALSE)
write_rds("data/data.rds")
```

<!-- End code -->

In this study, we analyzed data from a large open dataset on PowerWash Simulator (PWS) play and psychological experiences [@vuorreIntensiveLongitudinalDataset2023]. The data was collected in a research edition of PWS that recorded gameplay events, game status records, and responses to psychological survey items. We developed the research edition of PWS in collaboration with PWS's developer, FuturLab, who made it freely available on Steam to anyone who owned the original game (£19.99 on 2023-09-20). From the players' perspective, the research edition was nearly identical to that of the main game with the addition of in-game pop-ups that inquired about psychological states during play.

### PowerWash Simulator

PWS is a first-person simulation game developed by FuturLab. In the game, players run a small power washing business and take jobs from a variety of clients in different locations in the form of levels. The core mechanic of PWS is aiming and using a pressure washer to remove dirt from various objects and levels, ranging from Ferris wheels to skateparks. Progression happens sequentially through a career mode in which the player earns credits for cleaning objects and completing cleaning jobs. These credits can be used to upgrade the pressure washer to increase its range and effectiveness, as well as to purchase cosmetic modifications for the washer or avatar. The game offers a multiplayer mode which was disabled in the research version.

Critically, in addition to regular gameplay, the research edition surfaced psychological survey items to the player during play sessions. These survey items were integrated into the game as pop-ups using the existing in-game character dialogue system and delivered by a newly created character called "The Researchers" making them both conversational and part of the game lore, ensuring minimal disruption to the play experience. The maximum number of questions per hour was six, with a window of at least five minutes in between pop-ups. In addition, at the beginning of each play session, at player login, there was a 10% probability that the player was asked a question about their mood before starting play. Furthermore, players were also given the option to self-report mood in the main menu once every 30 minutes, but we excluded those menu reports in this manuscript.

### Participants

```{r}
#| label: demographics

demo <- read_csv(
  "data/demographics.csv",
  col_select = c(pid, country, gender, age)
) |>
  filter(pid %in% unique(dat$pid))

age <- median_qi(demo, age, .width = .8, na.rm = TRUE)

age <- str_glue("{age[1]} ({age[2]}, {age[3]}; 1st and 9th deciles)")

gender <- count(demo, gender, sort = TRUE) |>
  mutate(p = n / sum(n)) |>
  mutate(x = str_glue("{gender} ({number2(n)}, {percent2(p)})")) |>
  slice(1:4) |>
  pull(x) |>
  paste(collapse = ", ")

n_country <- length(unique(demo$country[!is.na(demo$country)]))
country <- count(demo, country, sort = TRUE) |>
  mutate(p = n / sum(n)) |>
  mutate(x = str_glue("{country} ({number2(n)}, {percent2(p)})")) |>
  slice(1:4) |>
  pull(x) |>
  knitr::combine_words()
```

The characteristics of the full sample of 11,080 players in the PWS dataset are described in @vuorreIntensiveLongitudinalDataset2023; here, we describe the subset of data relevant to our questions (see Data analysis below). All participants were over 18 years old, provided informed consent, answered at least one mood question, and did not request their data to be deleted. The median age was `r age`, and the four most frequent gender responses were `r gender`. Participants played in `r n_country` countries, with the `r country` being the most represented. Recruitment happened in multiple waves through multiple avenues inside and outside of the game [@vuorreIntensiveLongitudinalDataset2023]. Study participation was incentivized through cosmetic in-game rewards (e.g. item skins). For every 12 questions answered, players could unlock a reward, of which five were available. These rewards could only be unlocked in the research version but were usable in both the research and main versions of PWS.

The study procedures were granted ethical approval by Oxford University's Central University Research Ethics Committee (SSH_OII_CIA_21_011).

### Measures

We measured mood with a single item: "How are you feeling right now?" [@killingsworthWanderingMindUnhappy2010]. Participants responded using a visual analogue scale (VAS) with endpoints "Very bad" and "Very good" that recorded 1000 possible values, which we rescaled to the unit interval (0-1) for this study. Consequently, our results can also be interpreted on the "[proportion] of maximum possible" scale [POMP; @cohenProblemUnitsCircumstance1999]. While well-being is often studied with multi-item scales to differentiate between dimensions of positive and negative affect, the frequent probing of mental states in this study required a minimally intrusive instrument that would interrupt the participants' play experience as little as possible. Moreover, such single-item assessments have previously been validated and are recommended for intensive longitudinal studies [@songExaminingConcurrentPredictive2022].

### Data analysis

For the analyses reported here, we used a subset of the data in @vuorreIntensiveLongitudinalDataset2023 that was relevant to our questions. The full dataset contains `r number2(filter(dat_sum, when=="raw")$n_obs)` in-game survey responses, but here we ignored the enjoyment, focus, autonomy, competence, and immersion items and focused on the `r number2(filter(dat_sum, when=="mood")$n_obs)` mood responses from `r number2(filter(dat_sum, when=="mood")$n_session)` sessions and `r number2(filter(dat_sum, when=="mood")$n_pid)` players. We then excluded sessions longer than 5 hours in duration (`r percent2(1 - filter(dat_sum, when=="5h")$n_session / filter(dat_sum, when=="mood")$n_session)`) and dropped all responses with missing values (`r percent2(1 - filter(dat_sum, when=="non-na")$n_session / filter(dat_sum, when=="5h")$n_session)`). We made these decisions to reduce the complexity of our anticipated models, and under the belief that very long sessions are likely to be qualitatively different, and very rare, compared to typically shorter sessions. Our final dataset consisted of `r number2(filter(dat_sum, when=="non-na")$n_obs)` mood responses from `r number2(filter(dat_sum, when=="non-na")$n_session)` sessions and `r number2(filter(dat_sum, when=="non-na")$n_pid)` players.

Our first and main research question concerned the difference between players' moods at the beginning of each session (pre-play) and during the subsequent play session (during play). This contrast does not represent a causal hypothesis: Players could begin (and end) their play sessions for whatever reason, and these reasons are likely to confound the pre – during contrast. For example, a player might come home after a stressful day at work and then play PWS. Coming home from a stressful work environment might then cause the person to both (1) choose to play, and (2) experience an elevated mood, in which case we would be in error if we attributed play itself the position of causal antecedent of any potential mood consequences. Generally, reasons for starting to play are likely to contribute to the pre-during contrast and we are unable to disentangle those from any changes specifically caused by play.

We estimated this contrast within a three-level hierarchical regression model that nested observations within sessions, and sessions within players. We specified a censored Gaussian model of mood because a VAS necessarily limits response options at the lower and upper ends. Ignoring censoring would leave the contrast susceptible to ceiling or floor effects and might confound changes in the mood distribution's location with changes in its scale. We then modelled mean mood on an intercept and contrast coded time (pre-play: -0.5; during play: 0.5) and allowed both parameters to vary randomly across players. In addition, we modelled random intercepts over player sessions. Although equal residual variances across people in natural observation seem unlikely, we estimated only one residual deviation parameter to limit model complexity.

We analyzed the data with R and used the brms package to estimate, via Stan's HMC sampling algorithm, and post-process the models [@burknerBrmsPackageBayesian2017; @rcoreteamLanguageEnvironmentStatistical2023; @standevelopmentteamStanModelingLanguage2021]. We drew `r number2(N_ITER*4/2)` samples from the model's posterior distribution using brms' default prior distributions on all parameters and used numerical and graphical checks to ensure model convergence and adequacy.

```{r}
#| label: model-1-brms

model <- bf(
  mood | cens(cl) ~
    1 + post +
    (1 + post | pid) +
    (1 | ps)
) +
  gaussian()

fit1 <- brm(
  model,
  data = dat,
  silent = 0,
  iter = N_ITER,
  control = list(adapt_delta = .95),
  file = "models/brm-prepost-pid-ps-censored"
)
```

```{r}
#| label: fig-convergence
#| fig.cap: Convergence diagnostic plot showing bivariate scatterplots of the model's population-level parameters' posterior draws.
#| eval: false

pairs(
  fit1,
  variable = c("b_", "sd_", "cor_"),
  regex = TRUE,
  off_diag_args = list(size = .33, shape = 1)
)
```

## Results

```{r}
#| label: fig-descriptives
#| fig-height: 2.2
#| fig-width: 7
#| include: false
#| fig.cap: 'A. Histogram of session durations (note log10 x-axis). B. Summary of how many sessions each participant completed. C. Histogram of all mood ratings.'

p0 <- dat |>
  ggplot() +
  scale_x_continuous(
    expand = expansion(c(0.01, 0.01)),
  ) +
  scale_y_continuous(
    expand = expansion(c(0.001, 0.05)),
  ) +
  geom_histogram(
    col = "white",
    bins = 50,
    linewidth = .25,
    boundary = 0
  )

dat_session_durations <- dat |>
  filter(hours == max(hours), .by = c(pid, session))

dat_session_counts <- dat |>
  distinct(pid, session) |>
  count(pid, name = "sessions")

p_durations <- p0 %+%
  dat_session_durations +
  aes(hours) +
  labs(x = "Session duration (hours)", y = "Sessions")

p_sessions <- p0 %+%
  dat_session_counts +
  coord_cartesian(xlim = c(0, 75)) +
  aes(sessions) +
  labs(x = "Sessions", y = "Participants")

p_mood <- p0 +
  aes(mood) +
  labs(x = "Mood", y = "Responses")

(p_durations | p_sessions | p_mood) +
  plot_annotation(tag_levels = "A")
```

```{r}
#| include: false
tmp <- list(
  dat_session_durations$hours,
  dat_session_counts$sessions,
  dat$mood,
  dat$mood[dat$post == 0],
  dat$mood[dat$post == 1]
) |>
  map(
    ~ median_qi(.x, .width = 0.8, na.rm = TRUE)
  ) |>
  map(
    ~ mutate(
      .x,
      across(
        c(y, ymin, ymax),
        ~ number2(., .01)
      )
    ) |>
      str_glue_data("{y} [{ymin}, {ymax}]")
  )
tmp
```

The median session duration was `r tmp[[1]]` hours [10 and 90 percentiles]; the median player contributed data from `r tmp[[2]]` sessions, and the median mood was `r tmp[[3]]` (pre-session: `r tmp[[4]]`, during play: `r tmp[[5]]`). We illustrate these basic features of the data in Figure \@ref(fig:fig-data).

### RQ1: Mood changes from pre- to during play

```{r}
#| label: fig-data
#| include: true
#| fig-width: 8
#| fig-height: 5
#| fig.env: "figure*"
#| fig.cap: "A. Scatterplots of three participants' (rows) mood responses (pre-play: red; during play: blue) over eight sessions' (columns) durations. B. Histograms of session-mean (C. person-mean) moods before (top) and during (bottom) play sessions. D. Differences in session-mean (E. player-mean) mood differences (during session - pre-play). F. Scatterplot of person-mean mood reports at the beginning (x-axis) and during gameplay sessions (y-axis). Identity line is shown in green, and an exploratory GAM regression line is shown in blue."

# Plot: Example raw sessions
set.seed(99)
dat_example_sessions <- dat |>
  add_count(pid, session, name = "obs_per_session") |>
  filter(obs_per_session >= 4) |>
  mutate(session = as.numeric(as.factor(session)), .by = pid) |>
  mutate(session_per_person = length(unique(session)), .by = pid)

# default
min_sessions <- 8
n_filtered <- dat_example_sessions |>
  distinct(pid, session_per_person) |>
  filter(session_per_person >= 8) |>
  nrow()

# pick anyone if there's not 3 pids with >= 4 sessions
if (n_filtered < 3) min_sessions <- 0
dat_example_sessions <- dat_example_sessions |>
  filter(session_per_person >= min_sessions) |>
  arrange(pid, session) |>
  filter(pid %in% sample(unique(pid), 3)) |>
  mutate(
    Session = str_glue("Session {session}"), 
    Person = str_glue("Person {fct_anon(pid)}")
  )

p_mood_example <- dat_example_sessions |>
  filter(session <= 8) |>
  ggplot() +
  aes(hours, mood, col = post) +
  scale_color_brewer(
    "Pre-session measure",
    palette = "Set1",
    aesthetics = c("color", "fill")
  ) +
  scale_x_continuous(
    "Session duration (hours)"
  ) +
  scale_y_continuous(
    "Mood",
    limits = c(0, 1),
  ) +
  geom_point(size = 1.5, alpha = 1) +
  facet_grid(
    rows = vars(Person),
    cols = vars(Session),
    scales = "fixed"
  ) +
  theme(
    legend.position = "none",
    strip.text.y = element_blank()
  )

# Plot: Raw mood histograms at pre and during play
p_mood_prepost_raw <- dat |>
  mutate(post = factor(post, labels = c("Pre-play", "During play"))) |>
  ggplot(aes(mood, fill = post)) +
  scale_color_brewer(
    "Pre-session measure",
    palette = "Set1",
    aesthetics = c("color", "fill")
  ) +
  geom_histogram(bins = 30, col = "white") +
  scale_y_continuous(
    "Observations",
    expand = expansion(c(0.001, 0.1)),
    breaks = scales::extended_breaks(5)
  ) +
  scale_x_continuous(
    "Mood",
    expand = expansion(c(0.001))
  ) +
  facet_wrap("post", ncol = 1, scales = "free_y") +
  theme(legend.position = "none")

# Plot: Person-session-mean mood histograms at pre and during play
p_mood_prepost_sessions <- p_mood_prepost_raw %+%
  summarise(
    p_mood_prepost_raw$data,
    mood = mean(mood, na.rm = TRUE), .by = c(post, pid, ps)
  ) +
  scale_y_continuous(
    "Sessions",
    expand = expansion(c(0.001, 0.1)),
    breaks = scales::extended_breaks(5)
  )

# Plot: Person-mean mood histograms at pre and during play
p_mood_prepost_players <- p_mood_prepost_sessions %+%
  summarise(
    p_mood_prepost_raw$data,
    mood = mean(mood, na.rm = TRUE), .by = c(post, pid)
  ) +
  scale_y_continuous(
    "Players",
    expand = expansion(c(0.001, 0.1)),
    breaks = scales::extended_breaks(5)
  )

# Plot: Difference histogram (sessions)
p_mood_difference_sessions <- p_mood_prepost_sessions$data |>
  pivot_wider(names_from = post, values_from = mood) |>
  mutate(Difference = `During play` - `Pre-play`) |>
  drop_na(Difference) |>
  ggplot(aes(Difference)) +
  geom_histogram(bins = 50, col = "white") +
  geom_vline(xintercept = 0, linewidth = .5, col = "#2ca25f") +
  scale_y_continuous(
    "Sessions",
    expand = expansion(c(0.001, 0.1))
  ) +
  scale_x_continuous(
    expand = expansion(c(0.001))
  ) +
  coord_cartesian(xlim = c(-.4, .6))

# Plot: Difference histogram (players)
p_mood_difference_players <- p_mood_prepost_players$data |>
  pivot_wider(names_from = post, values_from = mood) |>
  mutate(Difference = `During play` - `Pre-play`) |>
  drop_na(Difference) |>
  ggplot(aes(Difference)) +
  geom_histogram(bins = 50, col = "white") +
  geom_vline(xintercept = 0, linewidth = .5, col = "#2ca25f") +
  scale_y_continuous(
    "Players",
    expand = expansion(c(0.001, 0.1))
  ) +
  scale_x_continuous(
    expand = expansion(c(0.001))
  ) +
  coord_cartesian(xlim = c(-.4, .6))

p_mood_difference <- (
  (p_mood_difference_sessions +
     theme(
       axis.title.x = element_blank(),
       axis.text.x = element_blank(),
       axis.ticks.x = element_blank()
     )) /
    p_mood_difference_players
)

p_mood_biscatter_sessions <- p_mood_difference_sessions$data |>
  ggplot(aes(`Pre-play`, `During play`)) +
  scale_x_continuous(
    expand = expansion(0.01)
  ) +
  scale_y_continuous(
    expand = expansion(0.01)
  ) +
  geom_point(
    alpha = .2, size = 0.33, shape = 1
  ) +
  geom_abline(linewidth = .5, col = "#2ca25f") +
  geom_smooth(
    method = "gam",
    se = FALSE,
    linewidth = .75,
    col = "dodgerblue"
  ) +
  theme(aspect.ratio = 1)

p_mood_biscatter_players <- p_mood_difference_players$data |>
  ggplot(aes(`Pre-play`, `During play`)) +
  scale_x_continuous(
    expand = expansion(0.01)
  ) +
  scale_y_continuous(
    expand = expansion(0.01)
  ) +
  geom_point(
    alpha = .2, size = 0.33, shape = 1
  ) +
  geom_abline(linewidth = .5, col = "#2ca25f") +
  geom_smooth(
    method = "gam",
    se = FALSE,
    linewidth = .75,
    col = "dodgerblue"
  ) +
  theme(aspect.ratio = 1)

p_mood_example /
  (
    p_mood_prepost_sessions |
      p_mood_prepost_players |
      p_mood_difference |
      p_mood_biscatter_players
  ) +
  plot_layout(heights = c(5, 5)) +
  plot_annotation(tag_levels = "A")
```

We first focused on our primary research question: To what extent do PWS players' mood change from pre-play to during play? We visualized the relevant data in Figure \@ref(fig:fig-data): Panel A shows mood responses from three example participants' first eight sessions of play. Figure \@ref(fig:fig-data) B (C) then shows histograms of all sessions' (players') aggregated pre- and during play moods to facilitate visual comparison of the raw data. We show the differences in these aggregated moods in Figure \@ref(fig:fig-data) D (sessions) and E (players). Moreover, Figure \@ref(fig:fig-data) F shows the difference in player-mean pre- and during play moods across different values of pre-play moods. Overall, these figures suggested small increases in mood from pre- to during play, but also that there were broad distributions of this difference over sessions and players, and that the difference was greater for lower pre-play moods (Panel F).

```{r}
#| label: tbl-avg
#| include: true
#| tbl-cap: Summaries of key population-level estimates

fit_tbl <- as_draws_df(
  fit1,
  c("b_", "sd_", "sigma"),
  regex = TRUE
) |>
  transmute(
    `Pre-play` = b_Intercept + b_post1 * -0.5,
    `During play` = b_Intercept + b_post1 * 0.5,
    Difference = b_post1,
    `Difference (scaled)` = Difference /
      (sd_pid__Intercept + sd_pid__post1 + sd_ps__Intercept + sigma),
    `(SD) Difference` = sd_pid__post1,
    `Positive shifts` = pnorm(0, b_post1, sd_pid__post1, lower.tail = FALSE)
  ) |>
  summarise_draws(
    mean = ~ mean(.),
    ~ quantile2(., probs = c(.025, .975))
  ) |>
  mutate(
    res = str_glue("{number2(mean, .001)} [{number2(q2.5, .001)}, {number2(q97.5, .001)}]"),
    resp = str_glue("{percent2(mean, .1)} [{percent2(q2.5, .1)}, {percent2(q97.5, .1)}]"),
  )

# Show this one as a percentage
fit_tbl[6, "res"] <- fit_tbl[6, "resp"]

fit_tbl |> 
  mutate(Variable = variable, Estimate = res, .keep = "none") |> 
  papaja::apa_table(
    span_text_columns = FALSE, 
    caption = "Summaries of the hierarchical model's key population-level estimates.",
    note = "Numbers indicate posterior means and 95%CIs. Difference (scaled) is the standardized during play--pre-play difference."
  )
```

We then turned to the model's results regarding (differences) in players' moods. They confirmed the visual impressions described above: Table \@ref(tab:tbl-avg) indicates that the average PWS player experiences a `r filter(fit_tbl, variable=="Difference")$res` unit increase in mood during PWS play, on a VAS from 0 to 1. We also interpreted this difference in a different light by dividing it by the total random variation estimated by the model. This standardized pre-play – during play contrast was `r filter(fit_tbl, variable=="Difference (scaled)")$res`.

### RQ2: Heterogeneity in mood changes

Although the above quantities indicated that the average player's mood did increase from before to during play, they say little about the robustness of this difference in the population of similar players. We therefore next turned to our second research question: How heterogeneous are mood shifts in the population of similar PWS players? First, the standard deviation of the person-specific mood differences was `r filter(fit_tbl, variable=="(SD) Difference")$res`. In comparison to the average person's estimated difference (`r filter(fit_tbl, variable=="Difference")$res`), that quantity indicated a moderate degree of heterogeneity between individuals. More specifically, the model-predicted proportion of individuals expected to experience positive changes from pre- to during play was `r filter(fit_tbl, variable=="Positive shifts")$resp`.

In sum, the results from our model contrasting pre- and during play moods indicated small increases in mood during play, and that those changes were robust across people.

### RQ3: Time course of mood changes during play

The above analysis provides an easily interpretable contrast between during-play moods and moods just before play. However, it does not address the time course of moods *within* the sessions. We therefore next turned to our third question: How do (changes in) players' moods evolve during gameplay sessions? To answer, we used time as a continuous predictor and allowed mood changes during sessions to be non-linear by estimating a piecewise cubic spline with 4 degrees of freedom. Just like the main model, this was a three-level hierarchical model, with random intercepts at the session and participant levels, and random participant slopes for each piece of the spline. Moreover, in a separate model, we also examined how within-session change related to mood before play by including pretest values as a covariate and modeled the hour-by-pretest continuous interaction. We also modeled pretest mood with a cubic spline to allow the relationship to be non-linear.

```{r}
#| label: model-2-3-lmer

fit2 <- fit_cached(
  "models/lmm-ns4.Rds",
  lmer(
    mood ~ ns4(hours) + (1 | ps) + (1 + ns4(hours) | pid),
    data = dat
  )
)

dat_pre <- dat |>
  filter(
    # Session has a wellbeing measure at time = 0
    hours[1] == 0,
    .by = c(pid, session)
  ) |>
  mutate(
    pre = mood[1],
    .by = c(pid, session)
  ) |>
  filter(
    hours > 0
  ) |>
  droplevels()

fit3 <- fit_cached(
  "models/lmm-pre-interaction.Rds",
  lmer(
    mood ~ ns4(hours) * ns(pre, 5) + (1 + ns4(hours) | pid) + (1 | ps),
    data = dat_pre
  )
)
```

The main model without an interaction with pre-play mood included all mood responses, sessions, and participants as above. However, the interaction model required each session to have a pre-play mood measure, which led to `r number2(lme4::ngrps(fit3)["pid"])` players, `r number2(lme4::ngrps(fit3)["ps"])` sessions, and `r number2(nobs(fit3))` observations included in that model.

We chose not to use model censoring in these models due to the increased computational cost. However, we performed sensitivity analyses with and without censoring on a reduced data set (1000 participants), which indicated nearly identical results. At worst, ignoring censoring resulted in slightly different intercepts; that is, the whole curve was shifted up or down.

```{r}
#| label: fig-ct1
#| include: true
#| fig.cap: "Estimated (changes) in mood as a function of session duration. Top: Average mood during a gaming session. Bottom: Change in mood during a session compared to mood at the beginning of a session. Gray ribbons indicate 95\\% confidence bands. We truncated the x-axis at three hours for this figure."
#| fig-height: 3
#| fig-width: 5

XMAX <- 3
XOUT <- 100
hours <- seq(0, XMAX, length.out = XOUT)
emm2_mood <- emmeans(
  fit2,
  ~hours,
  at = list(hours = hours),
  lmer.df = "asymptotic"
)
